Flexible Scientific Data Management for Plant Phenomics Research

### 2013-03-11 s4biodiv2013/PODD-S4BIODIV-Submission.tex
more cutbacks

Increasingly however, providers of scientific datasets are focusing on enhancing
their datasets using curated scientific concepts from ontologies. For example,
scientists have used the Gene Ontology \cite{Ashburner2000} to link well known concepts to
represent common elements across genomics datasets, while the Plant Ontology \cite{Avraham2008} 
allows the description of plant based datasets. If these annotations are
given using recognised URIs, and not as key-value string tuples, various RDF
documents can be merged automatically based on RDF semantics to discover novel
conclusions. Initially, different providers were not able to decide on URIs to
use in Linked Data. This resulted in various third parties such as Bio2RDF
\cite{Belleau2008}
creating RDF versions of scientific datasets so that the RDF documents could be
easily merged based on common URIs. However, recently more providers have given
authoritative URIs for each object in their dataset, making it easy to integrate
data from various locations.

\subsection{Redesign of the Phenomics Ontology Driven Data repository}
The PODD repository relies on semantic web technologies to
manage phenomics data and metadata. Although both ontologies and mappings are
essential, in PODD it was necessary to build the system with a relaxed
ontological vocabulary to allow scientists to generate datasets before they were
certain about the semantic meaning of their work. In addition, it was necessary
for PODD to allow scientists to continue to maintain projects containing curated
scientific concepts alongside raw experimental data. The PODD repository was 
redesigned based on an evaluation of the original software \cite{Li2010} that 
found it was not able to scale sufficiently well to suit the HRPPC needs. The 
major differences to the old software is that projects are no longer the only 
supported top object type, and projects are not stored in multiple parts, as 
that approach was not able to scale as was originally hypothesised.

The basic PODD project is structured as a tree, which, when used with
PlantScan\textsuperscript{\texttrademark}, represents the different components
of a scientific project as
top-level branches. These include a branch for raw data, along with separate
branches for results, analysis, and publications related to the project. In the
case of raw data, the semantics are not clear and cannot be defined by the
platforms collecting the data. Scientists are able to sparsely attach meaning
using horizontal links between different branches, using the OWL Open World
Assumption to allow extra semantic annotations to be attached in future. For
example, they may annotate images of a plant with a trait in only one case,
without annotating other images. These horizontal links are not constrained by
the basic scientific project structure, so they are able to entail novel
meanings independent of the ontology, including untested hypotheses, process
annotations and meaningful scientific results.

@@

OWL is used to determine whether projects are both internally consistent, with
all objects having an explicit RDF type, and whether they are consistent with
the ontologies that they import. The main points that PODD focuses on for
validation is determining that mappings are consistent with the basic PODD
predicates and classes, so that they can be used to render both static HTML
pages and provide HTML based editing facilities. For example, any OWL object
property that has been defined to link from image acquisition runs to images can
be used to define the purpose of an image. This may include any combination of
the defined properties, without requiring all of the properties to be used.

The parent object types and properties used by PODD to establish basic ground
rules for rendering and validation are defined in a small base schema ontology.
General scientific properties and phenotype specific properties are then defined
in larger extension ontologies as illustrated in Figure~\ref{poddontologies}.

>>>>

Increasingly however, providers of scientific datasets are focusing on enhancing
their datasets using curated scientific concepts from ontologies. For example,
scientists have used the Gene Ontology \cite{Ashburner2000} to link well known concepts to
represent common elements across genomics datasets, while the Plant Ontology \cite{Avraham2008} 
allows the description of plant based datasets. 
% If these annotations are
% given using recognised URIs, and not as key-value string tuples, various RDF
% documents can be merged automatically based on RDF semantics to discover novel
% conclusions. 
% Initially, different providers were not able to decide on URIs to
% use in Linked Data. This resulted in various third parties such as Bio2RDF
% \cite{Belleau2008}
% creating RDF versions of scientific datasets so that the RDF documents could be
% easily merged based on common URIs. However, recently more providers have given
% authoritative URIs for each object in their dataset, making it easy to integrate
% data from various locations.

\subsection{Redesign of the Phenomics Ontology Driven Data repository}
The PODD repository relies on semantic web technologies to
manage phenomics data and metadata. Although both ontologies and mappings are
essential, in PODD it was necessary to build the system with a relaxed
ontological vocabulary to allow scientists to generate datasets before they were
certain about the semantic meaning of their work. In addition, it was necessary
for PODD to allow scientists to continue to maintain projects containing curated
scientific concepts alongside raw experimental data. The PODD repository was 
redesigned based on an evaluation of the original software \cite{Li2010} that 
found it was not able to scale sufficiently well to suit the HRPPC needs. The 
major differences to the software implemented by \cite{Li2010} is that projects 
are no longer the only supported top object type, and projects are not stored 
in multiple parts, as that approach was not able to scale as was originally hypothesised.

The basic PODD project is structured as a tree, which, when used with
PlantScan\textsuperscript{\texttrademark}, represents the different components
of a scientific project as
top-level branches. These include a branch for raw data, along with separate
branches for results, analysis, and publications related to the project. In the
case of raw data, the semantics are not clear and cannot be defined by the
platforms collecting the data. 
% Scientists are able to sparsely attach meaning
% using horizontal links between different branches, using the OWL Open World
% Assumption to allow extra semantic annotations to be attached in future. 
For example, they may annotate images of a plant with a trait in only one case,
without annotating other images. These horizontal links are not constrained by
the basic scientific project structure, so they are able to entail novel
meanings independent of the ontology, including untested hypotheses, process
annotations and meaningful scientific results.

@@

OWL is used to determine whether projects are both internally consistent, with
all objects having an explicit RDF type, and whether they are consistent with
the ontologies that they import. The main points that PODD focuses on for
validation is determining that mappings are consistent with the basic PODD
predicates and classes, so that they can be used to render both static HTML
pages and provide HTML based editing facilities. For example, any OWL object
property that has been defined to link from image acquisition runs to images can
be used to define the purpose of an image. This may include any combination of
the defined properties, without requiring all of the properties to be used.
% The parent object types and properties used by PODD to establish basic ground
% rules for rendering and validation are defined in a small base schema ontology.
General scientific properties and phenotype specific properties are defined
in larger extension ontologies as illustrated in Figure~\ref{poddontologies}.
